Compensated de Casteljau algorithm in \(K\) times the working precision

### 2018-05-16 doc/paper.tex
Adding cover letter and doing final pass of manuscript.

\usepackage[usenames, dvipsnames]{color}
\usepackage{hyperref}
\hypersetup{
  colorlinks=true,
  pdfinfo={
    CreationDate={D:20180425160342},
    ModDate={D:20180425160342},
  },
}

@@

To enable a bound on the \(K\)-order error \(\db{K}^{(0)}_0\), it's necessary
to understand the difference between the exact local errors \(\ell_{F, j}\)
and the computed equivalents \(\widehat{\ell}_{F, j}\). To do this, we define
\begin{equation}
\widetilde{\ell}_{F, j} \coloneqq \left|e_1\right| +
\cdots + \left|e_{5F - 2}\right| + \left|\rho \cdot
\cdb{F - 1}_j^{(k + 1)}\right|.
\end{equation}

@@

\begin{lemma}\label{lemma:k-order}
The \(K\)-order error \(\db{K}^{(0)}_0\) satisfies
\begin{equation}
  \left|\db{K - 1}^{(0)}_0 - \cdb{K - 1}^{(0)}_0\right| =
  \left|\db{K}^{(0)}_0\right| \leq
  \left[\left(3^{K} \binom{n}{K} + \bigO{n^{K - 1}}\right)
  \mach^{K} + \bigO{\mach^{K + 1}}\right] \cdot \widetilde{p}(s).
\end{equation}
\end{lemma}

@@

All experiments were performed in IEEE-754 double precision.
As in \cite{Jiang2010}, we consider the evaluation in the neighborhood
of the multiple root of \(p(s) = (s - 1)\left(s - \frac{3}{4}\right)^7\),
written in Bernstein form.

@@

Figure~\ref{fig:smooth-drawing} shows the evaluation of \(p(s)\) at
the 401 equally spaced\footnote{It's worth noting that \(0.1\) cannot
be represented exactly in IEEE-754 double precision (or any binary
arithmetic for that matter). Hence (most of) the points of the form
\(a + b \cdot 10^{-c}\) can only be approximately represented.} points
\(\left\{\frac{3}{4} + j \frac{10^{-7}}{2}\right\}_{j=-200}^{200}\)
with \texttt{DeCasteljau} (Algorithm~\ref{alg:de-casteljau}),
\texttt{CompDeCasteljau} (Algorithm~\ref{alg:comp-de-casteljau})
and \texttt{CompDeCasteljau3} (Algorithm~\ref{alg:k-comp-de-casteljau}
with \(K = 3\)). We see that \texttt{DeCasteljau} fails to get the
magnitude correct, \texttt{CompDeCasteljau} has the right shape but
lots of noise and \texttt{CompDeCasteljau3} is able to smoothly evaluate
the function. This is in contrast to a similar figure in \cite{Jiang2010},
where the plot was smooth for the 400 equally spaced points
\(\left\{\frac{3}{4} + \frac{10^{-4}}{2} \frac{2j - 399}{399}
\right\}_{j=0}^{399}\). The primary difference is that as the interval
shrinks by a factor of \(\approx \frac{10^{-4}}{10^{-7}} = 10^3\), the
condition number goes up by \(\approx 10^{21}\) and \texttt{CompDeCasteljau}
is no longer accurate.

@@

Figure~\ref{fig:compensated-k} shows the relative forward errors compared
against the condition number. To compute relative errors, each input and
coefficient is converted to a fraction (i.e. infinite precision) and
\(p(s)\) is computed exactly as a fraction, then
compared to the corresponding computed values. Similar tools are used to
\textbf{exactly} compute the condition number, though here we can rely
on the fact that \(\widetilde{p}(s) = (s - 1)
\left(\frac{s}{2} - \frac{3}{4}\right)^7\). Once the relative errors and
condition numbers are computed as fractions, they are rounded to the
nearest IEEE-754 double precision value. As in \cite{Jiang2010}, we use
values \(\left\{\frac{3}{4} - (1.3)^j\right\}_{j=-5}^{-90}\)\footnote{As with
\(0.1\), it's worth noting that \((1.3)^j\) can't be represented exactly in
IEEE-754 double precision. However, this geometric series still serves a
useful purpose since it continues to raise \(\cond{p, s}\) as \(j\) decreases
away from \(0\) and because it results in ``random'' changes in the bits of
\(0.75\) that are impacted by subtracting \((1.3)^j\).}. The curves for
\texttt{DeCasteljau} and \texttt{CompDeCasteljau} trace the same paths seen
in \cite{Jiang2010}. In particular, \texttt{CompDeCasteljau} has a relative
error that is \(\bigO{\mach}\) until \(\cond{p, s}\) reaches
\(1 / \mach\), at which point the relative error increases linearly with
the condition number until it becomes \(\bigO{1}\) when
\(\cond{p, s}\) reaches \(1 / \mach^2\).
Similarly, the relative error in \texttt{CompDeCasteljau3}
(Algorithm~\ref{alg:k-comp-de-casteljau} with \(K = 3\))
is \(\bigO{\mach}\) until \(\cond{p, s}\) reaches
\(1 / \mach^2\) at which point the relative error increases linearly
to \(\bigO{1}\) when \(\cond{p, s}\) reaches \(1 / \mach^3\)
and the relative error in \texttt{CompDeCasteljau4}
(Algorithm~\ref{alg:k-comp-de-casteljau} with \(K = 4\))
is \(\bigO{\mach}\) until \(\cond{p, s}\) reaches
\(1 / \mach^3\) at which point the relative error increases linearly
to \(\bigO{1}\) when \(\cond{p, s}\) reaches \(1 / \mach^4\).

>>>>

\usepackage[usenames, dvipsnames]{color}
\usepackage{hyperref}
\hypersetup{
  colorlinks=true,
  pdfinfo={
    CreationDate={D:20180516090844},
    ModDate={D:20180516090844},
  },
}

@@

To enable a bound on the \(K\) order error \(\db{K}^{(0)}_0\), it's necessary
to understand the difference between the exact local errors \(\ell_{F, j}\)
and the computed equivalents \(\widehat{\ell}_{F, j}\). To do this, we define
\begin{equation}
\widetilde{\ell}_{F, j} \coloneqq \left|e_1\right| +
\cdots + \left|e_{5F - 2}\right| + \left|\rho \cdot
\cdb{F - 1}_j^{(k + 1)}\right|.
\end{equation}

@@

\begin{lemma}\label{lemma:k-order}
The \(K\) order error \(\db{K}^{(0)}_0\) satisfies
\begin{equation}
  \left|\db{K - 1}^{(0)}_0 - \cdb{K - 1}^{(0)}_0\right| =
  \left|\db{K}^{(0)}_0\right| \leq
  \left[\left(3^{K} \binom{n}{K} + \bigO{n^{K - 1}}\right)
  \mach^{K} + \bigO{\mach^{K + 1}}\right] \cdot \widetilde{p}(s).
\end{equation}
\end{lemma}

@@

@@

All experiments were performed in IEEE-754 double precision.
As in \cite{Jiang2010}, we consider the evaluation in the neighborhood
of the multiple root of \(p(s) = (s - 1)\left(s - 3/4\right)^7\),
written in Bernstein form.
Figure~\ref{fig:smooth-drawing} shows the evaluation of \(p(s)\) at
the 401 equally spaced\footnote{It's worth noting that \(0.1\) cannot
be represented exactly in IEEE-754 double precision (or any binary
arithmetic for that matter). Hence (most of) the points of the form
\(a + b \cdot 10^{-c}\) can only be approximately represented.} points
\(\left\{\frac{3}{4} + j \frac{10^{-7}}{2}\right\}_{j=-200}^{200}\)
with \texttt{DeCasteljau} (Algorithm~\ref{alg:de-casteljau}),
\texttt{CompDeCasteljau} (Algorithm~\ref{alg:comp-de-casteljau})
and \texttt{CompDeCasteljau3} (Algorithm~\ref{alg:k-comp-de-casteljau}
with \(K = 3\)). We see that \texttt{DeCasteljau} fails to get the
magnitude correct, \texttt{CompDeCasteljau} has the right shape but
lots of noise and \texttt{CompDeCasteljau3} is able to smoothly evaluate
the function. This is in contrast to a similar figure in \cite{Jiang2010},
where the plot was smooth for the 400 equally spaced points
\(\left\{\frac{3}{4} + \frac{10^{-4}}{2} \frac{2j - 399}{399}
\right\}_{j=0}^{399}\). The primary difference is that as the interval
shrinks by a factor of \(\approx \frac{10^{-4}}{10^{-7}} = 10^3\), the
condition number goes up by \(\approx 10^{21}\) and \texttt{CompDeCasteljau}
is no longer accurate.

@@

Figure~\ref{fig:compensated-k} shows the relative forward errors compared
against the condition number. To compute relative errors, each input and
coefficient is converted to a fraction (i.e. infinite precision) and
\(p(s)\) is computed exactly as a fraction, then
compared to the corresponding computed values. Similar tools are used to
\textbf{exactly} compute the condition number, though here we can rely
on the fact that \(\widetilde{p}(s) = (s - 1)
\left(s/2 - 3/4\right)^7\). Once the relative errors and
condition numbers are computed as fractions, they are rounded to the
nearest IEEE-754 double precision value. As in \cite{Jiang2010}, we use
values \(\left\{\frac{3}{4} - (1.3)^j\right\}_{j=-5}^{-90}\)\footnote{As with
\(0.1\), it's worth noting that \((1.3)^j\) can't be represented exactly in
IEEE-754 double precision. However, this geometric series still serves a
useful purpose since it continues to raise \(\cond{p, s}\) as \(j\) decreases
away from \(0\) and because it results in ``random'' changes in the bits of
\(0.75\) that are impacted by subtracting \((1.3)^j\).}. The curves for
\texttt{DeCasteljau} and \texttt{CompDeCasteljau} trace the same paths seen
in \cite{Jiang2010}. In particular, \texttt{CompDeCasteljau} has a relative
error that is \(\bigO{\mach}\) until \(\cond{p, s}\) reaches
\(1 / \mach\), at which point the relative error increases linearly with
the condition number until it becomes \(\bigO{1}\) when
\(\cond{p, s}\) reaches \(1 / \mach^2\).
Similarly, the relative error in \texttt{CompDeCasteljau3}
(Algorithm~\ref{alg:k-comp-de-casteljau} with \(K = 3\))
is \(\bigO{\mach}\) until \(\cond{p, s}\) reaches
\(1 / \mach^2\) at which point the relative error increases linearly
to \(\bigO{1}\) when \(\cond{p, s}\) reaches \(1 / \mach^3\)
and the relative error in \texttt{CompDeCasteljau4}
(Algorithm~\ref{alg:k-comp-de-casteljau} with \(K = 4\))
is \(\bigO{\mach}\) until \(\cond{p, s}\) reaches
\(1 / \mach^3\) at which point the relative error increases linearly
to \(\bigO{1}\) when \(\cond{p, s}\) reaches \(1 / \mach^4\).
